\documentclass[conference]{IEEEtran}
\usepackage{cite}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{url}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{array}
\usepackage{multirow}
\usepackage{listings}
\lstset{
  basicstyle=\footnotesize\ttfamily,
  breaklines=true,
  breakatwhitespace=false,
  columns=flexible,
  keepspaces=true
}

\title{Dynamic Load Balancing for a Real-Time Analytics Dashboard:\\
A Comparative Evaluation of Round-Robin, Least-Connections,\\
and Latency-Aware Routing}

\author{
\IEEEauthorblockN{Ana Maria Floristean, Razvan-Constantin Croitoru}
\IEEEauthorblockA{Faculty of Computer Science\\
Alexandru Ioan Cuza University, Ia\c{s}i, Romania\\
anafloris23@gmail.com, croitorurazvan1@gmail.com}
}

\begin{document}

\maketitle

\begin{abstract}
Distributed systems that serve real-time analytics must spread incoming
requests across multiple backend workers without creating hotspots or
causing cascading failures. In this paper we design, implement, and
evaluate a fully containerised load-balancing testbed featuring three
worker services --- two fast workers and one intentionally degraded
worker with a 4$\times$ slowdown factor --- and a custom reverse proxy
supporting three routing policies: static round-robin, least-connections,
and a latency-aware strategy based on Exponentially Weighted Moving
Average (EWMA) scoring. We evaluate the system under four controlled
workloads: constant load, traffic spike, progressive stress to 300
virtual users, and deliberate worker failure injection.

Results demonstrate clear differentiation between algorithms. Under
constant moderate load (30\,VUs), the latency-aware algorithm routes
only 4.9\% of traffic to the slow worker (versus 33\% for round-robin),
reducing average latency by 30\% and increasing throughput by 19\%.
Under extreme stress (300\,VUs), both adaptive algorithms sustain over
229\,req/s with under 2\% errors, while round-robin achieves only
173\,req/s with 7.6\% errors and drives all three workers into an
unhealthy state. In the failure-injection scenario, the system detects
the crashed worker within one health-check interval and reroutes traffic
with just 0.01\% total failures (1 of 9252 requests). We additionally
document two implementation correctness issues --- EWMA contamination by
health-check latency and a concurrency race on the in-flight counter ---
whose presence caused all algorithms to behave identically, and whose
resolution was necessary to observe the expected differentiation.
\end{abstract}

\begin{IEEEkeywords}
load balancing, round-robin, least-connections, EWMA, latency-aware
routing, distributed systems, containerisation, FastAPI, k6, Docker
Compose, performance evaluation, fault tolerance
\end{IEEEkeywords}

% ---------------------------------------------------------------
\section{Introduction}
% ---------------------------------------------------------------

Modern web applications are typically deployed as a collection of
stateless backend services replicated across multiple workers to achieve
horizontal scalability. A \emph{load balancer} sits in front of these
workers and decides, for each incoming request, which worker should
handle it. The decision policy --- the \emph{routing algorithm} ---
directly affects user-perceived latency, system throughput, and
resilience to failures.

The simplest policy, round-robin (RR), distributes requests evenly in
order. This is optimal when all workers are identical and have the same
capacity. In practice, heterogeneity is common: workers run on different
hardware generations, experience garbage-collection pauses, share CPU
with co-located services, or become partially degraded due to software
bugs or resource exhaustion. Under these conditions, a static algorithm
wastes capacity by over-routing to slow workers.

Dynamic policies address this by incorporating real-time signals.
Least-connections (LC) uses the current request queue depth; latency-based
policies use measured response times. Both signal sources are imperfect:
connection counts do not distinguish between a fast worker with many
small requests and a slow worker with few large ones; response-time
estimates must be tracked carefully to avoid convergence to stale values.

\textbf{Research question.} \textit{Which routing strategy provides the
best trade-off between throughput, tail latency, and resilience in a
heterogeneous worker pool under variable workload and failure?}

To answer this, we built a reproducible containerised testbed and
conducted four controlled experiments. Our contributions are:
\begin{enumerate}
    \item A working implementation of three routing algorithms in a single
    async Python reverse proxy, including their correctness pitfalls.
    \item A quantitative comparison across four workload regimes on
    real hardware, with all metrics captured from a live Prometheus
    endpoint and k6 summary exports.
    \item An integrated specification connecting each system requirement
    to a concrete experimental measurement or functional test.
    \item Documentation of two implementation bugs that silenced
    algorithm differentiation and the fixes required to restore correct
    behaviour.
\end{enumerate}

% ---------------------------------------------------------------
\section{Background and Related Work}
% ---------------------------------------------------------------

\subsection{Load Balancing Algorithms}

Load balancing has been studied since the early days of distributed
computing \cite{tanenbaum}. The three canonical algorithms that form our
baseline are:

\textbf{Round-robin.} The simplest deterministic policy; each new request
is routed to the next worker in a cyclic list. It requires no state beyond
a counter and performs optimally when service times are uniform \cite{cardellini}.

\textbf{Least-connections.} Introduced to handle uneven service times,
it routes each request to the worker with the smallest number of active
(in-flight) connections. Theoretical analysis shows this is near-optimal
when service times follow an exponential distribution \cite{mitzenmacher2001}.
Practical implementations track an atomic in-flight counter per worker.

\textbf{Latency-based / weighted.} A wider family of algorithms that
use historical response-time measurements to infer worker capacity and
prefer faster workers. Pai et al.\ \cite{pai1998} demonstrate that
locality-aware distribution improves both throughput and cache hit rates
in cluster-based servers. Envoy Proxy \cite{envoy} implements an EWMA
variant in production.

\subsection{EWMA for Latency Estimation}

Exponentially weighted moving average smooths a noisy signal by giving
more weight to recent observations:
\begin{equation}
    S_t = \alpha \, x_t + (1-\alpha) \, S_{t-1}, \quad \alpha \in (0,1)
\end{equation}
A higher $\alpha$ reacts faster to changes but amplifies noise; a lower
$\alpha$ is more stable but slower to converge. Jacobson \cite{jacobson1988}
originally applied EWMA to TCP round-trip-time estimation. The same
technique has been adopted for load balancing by Envoy \cite{envoy},
where a separate EWMA per upstream tracks recent request latency.

\subsection{Health Checking and Failure Detection}

Active health checking --- periodically pinging each backend with a
lightweight request --- is the standard mechanism for failure detection
in reverse proxies such as NGINX \cite{nginx} and HAProxy \cite{haproxy}.
The trade-off between check frequency and the window of forwarded
failures is well understood \cite{tanenbaum}. Less commonly studied is
the interaction between health-check latency and routing-signal
estimation: if a probe is issued to an idle worker, the measured latency
reflects idle network RTT, not true request-processing capacity.

\subsection{Containerised Testbeds}

Docker Compose has become the standard tool for reproducible multi-service
experiments on a single host \cite{docker}. Kubernetes \cite{kubernetes}
extends this to multi-host clusters but adds scheduling complexity that
obscures the routing algorithm's effect. We use Docker Compose to isolate
the routing algorithm as the sole independent variable across runs.

% ---------------------------------------------------------------
\section{Problem Formulation}
% ---------------------------------------------------------------

\subsection{System Model}

Let $W = \{w_1, w_2, \ldots, w_n\}$ be a set of $n$ backend workers.
Each worker $w_i$ has a true service-time distribution $F_i$ for a given
request type. In our testbed, $n=3$; workers $w_1$ and $w_2$ are
homogeneous with mean service time $\mu_{\text{fast}} \approx 45$\,ms,
and $w_3$ is degraded with $\mu_{\text{slow}} \approx 180$\,ms. Worker
capacities are unknown to the load balancer a priori and must be
inferred from observations.

Requests arrive at the load balancer at rate $\lambda$ (VUs generating
requests with inter-arrival determined by response time and a fixed
sleep). The load balancer maintains a state vector
$\mathbf{s} = (\text{in\_flight}_i, \hat{\mu}_i, h_i)_{i=1}^{n}$,
where $\text{in\_flight}_i$ is the current number of outstanding
requests, $\hat{\mu}_i$ is the estimated mean service time (via EWMA),
and $h_i \in \{0,1\}$ is the health status.

\subsection{Routing Policies}

A routing policy $\pi : \mathbf{s} \rightarrow W$ maps the current
state to a worker selection. We evaluate three policies:

\begin{enumerate}
    \item \textbf{Round-Robin:} $\pi_{\text{RR}}(\mathbf{s}) = w_{(k \bmod n)+1}$,
    where $k$ increments atomically with each request and only healthy
    workers are counted.
    \item \textbf{Least-Connections:}
    $\pi_{\text{LC}}(\mathbf{s}) = \arg\min_{w_i : h_i=1}
    \text{in\_flight}_i$, with ties broken by $\hat{\mu}_i$.
    \item \textbf{Latency-Aware (EWMA):}
    $\pi_{\text{LA}}(\mathbf{s}) = \arg\min_{w_i : h_i=1}
    \bigl(\hat{\mu}_i + \beta \cdot \text{in\_flight}_i\bigr)$,
    where $\beta = 5$\,ms is a penalty coefficient.
\end{enumerate}

\subsection{Optimisation Objectives}

We evaluate each policy against three objectives, measured empirically:
\begin{itemize}
    \item \textbf{Throughput} $T$ (requests per second): higher is better.
    \item \textbf{Average and tail latency} ($\bar{L}$, $L_{95}$): lower is better.
    \item \textbf{Error rate} $\varepsilon$ (fraction of requests timing out
    or receiving 5xx): lower is better.
\end{itemize}

These objectives are partly in tension: maximising throughput under
overload requires accepting some increase in tail latency.

% ---------------------------------------------------------------
\section{Proposed System Design}
% ---------------------------------------------------------------

\subsection{Architecture Overview}

The system follows a four-tier architecture, with all services running
as Docker containers on a single host connected via a Docker bridge
network.

\begin{enumerate}
    \item \textbf{Workers} ($n=3$): FastAPI async HTTP services that
    simulate analytics event processing with configurable latency.
    \item \textbf{Load Balancer}: A custom FastAPI reverse proxy that
    implements the three routing policies, performs health checking,
    tracks EWMA latencies, and exposes Prometheus metrics.
    \item \textbf{Observability}: Prometheus scrapes
    \texttt{/lb/metrics} every 2\,s; Grafana provides real-time dashboards.
    \item \textbf{Load Generator}: k6 virtual users drive HTTP traffic
    according to configurable VU ramp profiles.
\end{enumerate}

A supplementary \emph{test-runner} service wraps k6 as a subprocess and
exposes an HTTP API so that experiments can be triggered from the
browser dashboard without SSH access.

\subsection{Worker Service Design}

Each worker exposes three endpoints:
\begin{itemize}
    \item \texttt{GET /api/v1/analytics/event?delay\_ms=\&cpu\_ms=} ---
    processes a synthetic analytics event. The handler first executes a
    synchronous busy-loop for \texttt{cpu\_ms} milliseconds (blocking the
    asyncio event loop), then awaits \texttt{asyncio.sleep(delay\_ms/1000)}.
    \item \texttt{GET /api/v1/health} --- returns 200 OK immediately;
    used by the health-check loop.
    \item \texttt{POST /api/v1/crash} --- calls \texttt{os.\_exit(1)},
    simulating an abrupt process failure for the failure-injection scenario.
\end{itemize}

Worker heterogeneity is introduced via the \texttt{SLOWDOWN\_FACTOR}
environment variable, which multiplies both \texttt{delay\_ms} and
\texttt{cpu\_ms} before processing. Worker-3 uses
\texttt{SLOWDOWN\_FACTOR=4}, yielding 4$\times$ longer response times
than workers 1 and 2.

\subsection{Load Balancer Design}

The load balancer is a FastAPI application with a single catch-all
route that proxies all non-management requests to a selected worker.
Per-worker state is maintained in a dataclass:

\begin{lstlisting}
@dataclass
class Worker:
    name: str
    base_url: str
    healthy: bool = True
    in_flight: int = 0
    ewma_latency_ms: float = 100.0
    total_requests: int = 0
    failed_requests: int = 0
\end{lstlisting}

The \texttt{select\_worker()} coroutine filters to healthy workers,
then applies the active routing policy. A background
\texttt{health\_loop()} coroutine concurrently pings all workers every
\texttt{HEALTH\_CHECK\_INTERVAL\_MS} (default 1000\,ms) with a timeout
of \texttt{HEALTH\_TIMEOUT\_MS} (default 800\,ms), updating the
\texttt{healthy} flag on failure or recovery.

% ---------------------------------------------------------------
\section{Implementation}
% ---------------------------------------------------------------

\subsection{Routing Algorithm Implementation}

The three policies share a common \texttt{select\_worker()} entry point
that dispatches based on the \texttt{ALGORITHM} environment variable.
Round-robin uses a module-level counter protected by
\texttt{asyncio.Lock} to prevent race conditions. Least-connections and
latency-aware use \texttt{min(candidates, key=...)} over the healthy
worker list.

\subsection{EWMA Update Correctness}

A critical implementation detail is that the EWMA \emph{must not} be
updated during health checks. Health probes are issued to potentially
idle workers and measure network RTT rather than request service time.
If health-check latency is folded into the routing EWMA, all workers
converge to similar low latency values during idle periods, eliminating
the signal that allows latency-aware routing to prefer fast workers.

The correct update location is inside \texttt{proxy\_to\_worker()},
after the upstream response is received:

\begin{lstlisting}
duration_ms = (time.perf_counter() - started) * 1000
worker.ewma_latency_ms = ewma(
    worker.ewma_latency_ms, duration_ms)
\end{lstlisting}

The \texttt{run\_health\_check()} function updates only
\texttt{worker.healthy} and \texttt{worker.last\_health\_check\_at}.

\subsection{In-Flight Counter Race Condition}

In Python's asyncio model, every \texttt{await} expression is a
\emph{yield point} at which the event loop may schedule another
coroutine. The in-flight counter must be incremented \emph{before} any
such yield point in the request-handling path. In particular,
\texttt{await request.body()} reads the request body and yields;
multiple concurrent coroutines that each call \texttt{select\_worker()}
before any increments the counter will all see \texttt{in\_flight=0}
for all workers and may route to the same worker simultaneously.

The fix is to increment the counter immediately after worker selection:

\begin{lstlisting}
worker = select_worker()   # select before awaits
worker.in_flight += 1      # increment immediately
worker.total_requests += 1
stats.total_requests += 1
body = await request.body()  # yield point (safe now)
\end{lstlisting}

\subsection{Experiment Script Correctness}

Each experiment scenario restarts the full Docker Compose stack with a
different \texttt{ALGORITHM} environment variable. The routing policy
is read by the load balancer at startup via \texttt{os.getenv()}. A
subtle issue was that the \texttt{docker compose --profile loadtest run}
command (which launches k6 as a one-off container) was initially invoked
\emph{without} the \texttt{ALGORITHM} variable prefix. Docker Compose
resolved the missing variable to its default (\texttt{round\_robin})
and recreated the load-balancer container, overwriting the intended
algorithm. This was fixed by prefixing every \texttt{docker compose}
invocation in the experiment loop with the appropriate variable:

\begin{lstlisting}
ALGORITHM="$algorithm" docker compose \
    --profile loadtest run --rm k6 run ...
\end{lstlisting}

\subsection{Observability and Dashboard}

The load balancer exposes a Prometheus text-format metrics endpoint at
\texttt{/lb/metrics} with per-worker gauges (\texttt{in\_flight},
\texttt{ewma\_latency\_ms}, \texttt{healthy}, \texttt{score}) and
system-wide counters (\texttt{lb\_requests\_total},
\texttt{lb\_failed\_requests\_total}, rolling p95/p99). A static HTML
dashboard at \texttt{/dashboard} polls these endpoints and additionally
allows operators to trigger k6 scenarios from the browser, streaming
live output via a long-polling endpoint (\texttt{/tests/output?since=N}).

% ---------------------------------------------------------------
\section{Experimental Methodology}
% ---------------------------------------------------------------

\subsection{Environment}

All experiments run on a personal home server equipped with an Intel
Core i5-7500T CPU (2.70\,GHz, 4 cores) and 15\,GB RAM, running Debian
Linux with Docker 27 and Docker Compose v2. Container-to-container
network latency is sub-millisecond and negligible compared to the
workload delays. Each experiment scenario is fully isolated: the entire
stack is stopped (\texttt{docker compose down}), rebuilt, and restarted
between runs to guarantee clean state and fresh worker EWMA values.
Results are collected from k6 summary JSON exports and from the load
balancer's \texttt{/lb/stats} and \texttt{/lb/workers} endpoints,
captured immediately after each k6 run completes.

\subsection{Workload Parameters}

All scenarios use \texttt{delay\_ms=40} and \texttt{cpu\_ms=5}. For
workers 1 and 2 this yields approximately 45\,ms per request (5\,ms
synchronous CPU work, 40\,ms async sleep). For worker-3
(\texttt{SLOWDOWN\_FACTOR=4}) this becomes approximately 180\,ms
(20\,ms CPU, 160\,ms sleep). The load balancer's request timeout is
2000\,ms; requests not completed within this window return a 502 error.

\subsection{Scenarios}

\textbf{S1 --- Constant Load.} 30 virtual users, 60\,s duration.
A steady-state baseline that allows EWMA values to converge and
measures the equilibrium traffic distribution and latency for each
algorithm.

\textbf{S2 --- Traffic Spike.} VUs ramp from 5 to 80 over 5\,s, hold
for 30\,s, then ramp back to 0. This tests algorithm responsiveness
when concurrency increases abruptly. The ramp is fast relative to
EWMA convergence, creating a transient period of less-informed routing.

\textbf{S3 --- Stress Test.} VUs ramp from 5 to 300 over 70\,s.
This progressively saturates the system and exposes the point at which
each algorithm transitions from stable to degraded operation. At 300\,VUs
the system is intentionally overloaded.

\textbf{S4 --- Failure Injection.} Constant 30 VUs using the
latency-aware algorithm for 60\,s; worker-2 is terminated via its
\texttt{/api/v1/crash} endpoint at $t=20$\,s. This measures detection
latency and the fraction of requests lost during the failure transition.

\subsection{Metrics and Collection}

Primary metrics are collected by k6 and include \texttt{http\_req\_duration}
(average and p95), \texttt{http\_reqs} (throughput), and
\texttt{http\_req\_failed} (error rate). Secondary metrics are collected
from the load balancer's JSON endpoints: per-worker \texttt{total\_requests},
\texttt{ewma\_latency\_ms}, and \texttt{healthy} status. All raw results
are stored as JSON files and parsed by a Python script
(\texttt{scripts/parse\_results.py}) to produce the tables in this paper.

% ---------------------------------------------------------------
\section{Results}
% ---------------------------------------------------------------

\subsection{S1: Constant Load}

Table~\ref{tab:constant} shows results for 30~VUs over 60\,s.

\begin{table}[h]
\caption{Constant Load Results (30 VUs, 60\,s)}
\label{tab:constant}
\centering
\begin{tabular}{@{}lccc@{}}
\toprule
Metric & RR & LC & LA \\
\midrule
Avg latency (ms)          & 107.1 & 79.7  & 74.4  \\
p95 latency (ms)          & 224.9 & 220.1 & 182.4 \\
Throughput (req/s)        & 143.9 & 165.7 & 170.7 \\
Error rate (\%)           & 0.0   & 0.0   & 0.0   \\
Traffic to worker-3 (\%) & 33.3  & 13.4  & 4.9   \\
Jain's fairness index $J$ & 1.000 & 0.847 & 0.733 \\
\bottomrule
\end{tabular}
\end{table}

All three algorithms achieve zero errors under moderate load. However,
traffic distributions differ markedly. Round-robin sends 33.3\% to each
worker as designed. Least-connections reduces worker-3's share to 13.4\%:
the slow worker's 180\,ms service time means it holds more in-flight
connections per unit time, making it the least-preferred choice.
Latency-aware is the most aggressive, sending only 4.9\% to worker-3
--- whose EWMA score ($\sim$200\,ms) is four times that of the fast
workers ($\sim$49\,ms).

Performance follows distribution: latency-aware achieves
74.4\,ms average latency, 30\% below round-robin's 107.1\,ms, with 19\%
higher throughput (170.7 vs.\ 143.9\,req/s). Least-connections is
intermediate (79.7\,ms, 165.7\,req/s).

\subsection{S2: Traffic Spike}

Table~\ref{tab:spike} shows spike results. The burst to 80\,VUs
doubles or triples throughput for adaptive algorithms while keeping
error rates at zero; round-robin lags significantly.

\begin{table}[h]
\caption{Spike Test Results (5 $\rightarrow$ 80 VUs)}
\label{tab:spike}
\centering
\begin{tabular}{@{}lccc@{}}
\toprule
Metric & RR & LC & LA \\
\midrule
Avg latency (ms)          & 392.1  & 182.0 & 172.2 \\
p95 latency (ms)          & 1401.8 & 485.3 & 362.3 \\
Throughput (req/s)        & 114.5  & 216.3 & 225.8 \\
Error rate (\%)           & 0.0    & 0.0   & 0.0   \\
Traffic to worker-3 (\%) & 33.3   & 15.4  & 12.0  \\
Jain's fairness index $J$ & 1.000 & 0.871 & 0.830 \\
\bottomrule
\end{tabular}
\end{table}

Round-robin's p95 of 1402\,ms indicates queuing at worker-3: fast
workers cannot process requests fast enough to avoid backlog when
one-third of traffic goes to a 4$\times$-slower worker. Adaptive
algorithms limit this to 12--15\%, keeping p95 under 500\,ms and
nearly doubling throughput (216--226 vs.\ 114\,req/s).

\subsection{S3: Stress Test}

Table~\ref{tab:stress} shows stress results at saturation.

\begin{table}[h]
\caption{Stress Test Results (5 $\rightarrow$ 300 VUs)}
\label{tab:stress}
\centering
\begin{tabular}{@{}lccc@{}}
\toprule
Metric & RR & LC & LA \\
\midrule
Avg latency (ms)          & 595.4  & 440.9  & 440.8  \\
p95 latency (ms)          & 2524.7 & 2601.7 & 2276.5 \\
Throughput (req/s)        & 173.3  & 231.7  & 228.7  \\
Error rate (\%)           & 7.59   & 1.63   & 0.80   \\
Traffic to worker-3 (\%) & 33.3   & 16.8   & 12.7   \\
Jain's fairness index $J$ & 1.000 & 0.891 & 0.839 \\
Workers healthy at end    & None   & All    & All    \\
\bottomrule
\end{tabular}
\end{table}

Round-robin saturates all three workers evenly, cascading into health
check failures (all workers marked unhealthy by test end) and 7.6\%
errors. Both adaptive algorithms keep all workers healthy and achieve
32\% higher throughput with 5--10$\times$ lower error rates. Latency-aware
has a slight edge over least-connections in error rate (0.80\% vs.\
1.63\%) and p95 (2277\,ms vs.\ 2602\,ms).

\subsection{S4: Failure Injection}

Table~\ref{tab:failure} shows failure-injection results using the
latency-aware algorithm.

\begin{table}[h]
\caption{Failure Injection Results (LA, 30 VUs, worker-2 crash at $t{=}20$\,s)}
\label{tab:failure}
\centering
\begin{tabular}{@{}lc@{}}
\toprule
Metric & Value \\
\midrule
Detection latency            & $<$1\,s (1 health-check interval) \\
k6 avg latency (full 60\,s)  & 94.1\,ms \\
k6 p95 latency               & 241.5\,ms \\
Throughput                   & 153.6\,req/s \\
Total errors (LB stats)      & 1 of 9252 requests (0.01\%) \\
Worker-1 (post-crash)        & 6\,462 requests, 0 failures \\
Worker-2 (pre-crash only)    & 1\,585 requests, 1 failure \\
Worker-3 (post-crash)        & 1\,205 requests, 0 failures \\
\bottomrule
\end{tabular}
\end{table}

Worker-2's crash is detected within one health-check interval. Only
one request fails (a 502 proxy error during the in-progress connection
at crash time). The latency-aware algorithm subsequently routes 84\%
of new traffic to worker-1 (EWMA $\approx$67\,ms) and 16\% to worker-3
(EWMA $\approx$215\,ms), maintaining 94\,ms average latency throughout.

\subsection{Resource Utilization (30 VUs, Constant Load)}

Table~\ref{tab:resources} shows container-level CPU and memory usage,
sampled via \texttt{docker stats} at mid-test (after EWMA convergence)
under the constant-load workload. CPU percentages are per-core (100\,\%
$=$ one fully used CPU core).

\begin{table}[h]
\caption{Resource Utilization per Container (30 VUs, constant load)}
\label{tab:resources}
\centering
\small
\begin{tabular}{@{}lrrrr@{}}
\toprule
Container & \multicolumn{2}{c}{CPU (\%)} & \multicolumn{2}{c}{Mem (MiB)} \\
\cmidrule(lr){2-3}\cmidrule(lr){4-5}
 & Idle & Under load & Idle & Under load \\
\midrule
\multicolumn{5}{l}{\textit{Round-Robin}} \\
Load balancer & 1.2 & 29.3 & 46.5 & 46.4 \\
Worker-1      & 0.3 & 26.3 & 32.9 & 32.9 \\
Worker-2      & 0.3 & 26.6 & 40.5 & 40.5 \\
Worker-3      & 0.2 & \textbf{98.5} & 32.8 & 32.9 \\
\midrule
\multicolumn{5}{l}{\textit{Least-Connections}} \\
Load balancer & 1.2 & 48.1 & 46.5 & 42.4 \\
Worker-1      & 0.3 & 45.4 & 32.9 & 32.9 \\
Worker-2      & 0.3 & 39.3 & 40.5 & 40.5 \\
Worker-3      & 0.2 & 42.0 & 32.8 & 32.9 \\
\midrule
\multicolumn{5}{l}{\textit{Latency-Aware}} \\
Load balancer & 1.2 & 45.7 & 46.5 & 44.4 \\
Worker-1      & 0.3 & 46.9 & 32.9 & 32.9 \\
Worker-2      & 0.3 & 47.9 & 40.5 & 40.5 \\
Worker-3      & 0.2 & \textbf{18.6} & 32.8 & 32.9 \\
\bottomrule
\end{tabular}
\end{table}

The most striking result is worker-3's CPU usage: under round-robin it
reaches 98.5\% (one saturated core), while under latency-aware it drops
to 18.6\%. This directly reflects the traffic share (33\% vs.\ 4.9\%)
and confirms that the algorithm prevents the CPU hotspot that round-robin
creates. Memory footprint is stable across all algorithms, with no growth
during the test, confirming the absence of memory leaks.

The load balancer itself consumes more CPU under adaptive algorithms
(48\% for LC/LA vs.\ 29\% for RR). However, adaptive algorithms also
handle more requests (166--171 vs.\ 144\,req/s), so the per-request LB
overhead is approximately 0.29\,\%/req (LC), 0.27\,\%/req (LA), and
0.20\,\%/req (RR) --- a 35--45\% overhead increase that is negligible
in absolute terms with $n=3$ workers.

% ---------------------------------------------------------------
\section{Integrated System Requirements and Conformance Specification}
% ---------------------------------------------------------------

\subsection{System Requirements}

Table~\ref{tab:requirements} lists the system requirements using formal
wording: \textbf{shall} for mandatory, \textbf{should} for recommended,
and \textbf{may} for optional capabilities.

\begin{table*}[t]
\caption{System Requirements and Conformance Specification}
\label{tab:requirements}
\centering
\small
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}p{0.5cm}p{5.5cm}p{1.0cm}p{2.8cm}p{3.5cm}@{}}
\toprule
\textbf{ID} & \textbf{Requirement} & \textbf{Pri.} & \textbf{Verification} & \textbf{Evidence} \\
\midrule
R1  & The system \textbf{shall} maintain at least three concurrent backend workers.
    & Must & Docker Compose \texttt{ps}
    & All four scenarios run with workers 1--3 active. \\
R2  & The load balancer \textbf{shall} implement round-robin routing as a selectable algorithm.
    & Must & Functional: uniform distribution check
    & S1 shows 33.3\%/33.3\%/33.3\% under RR. \\
R3  & The load balancer \textbf{shall} implement least-connections routing.
    & Must & Functional: worker-3 receives fewer requests than under RR
    & S1: LC routes 13.4\% to worker-3 vs.\ 33.3\% for RR. \\
R4  & The load balancer \textbf{shall} implement latency-aware (EWMA) routing.
    & Must & Functional: worker-3 share lower than LC
    & S1: LA routes 4.9\% to worker-3, lowest of all algorithms. \\
R5  & The system \textbf{shall} perform active health checks on all workers at most every 1\,s.
    & Must & Inspect health-check log timestamps
    & Health loop configured at 1000\,ms interval; verified in code. \\
R6  & Unhealthy workers \textbf{shall} be excluded from routing within 1\,s of failure detection.
    & Must & Failure injection (S4)
    & S4: worker-2 excluded after 1 interval; only 1 request failed. \\
R7  & The load balancer \textbf{shall} expose Prometheus-compatible metrics at \texttt{/lb/metrics}.
    & Must & \texttt{curl /lb/metrics} format check
    & Metrics file collected after each scenario; Grafana dashboard live. \\
R8  & The system \textbf{shall} support synthetic workload injection via k6.
    & Must & k6 summary JSON present after each run
    & 37 result files produced across all scenarios. \\
R9  & Workers \textbf{shall} support a configurable \texttt{SLOWDOWN\_FACTOR} to model heterogeneity.
    & Must & Compare EWMA: worker-3 $\approx$ 4$\times$ workers 1--2
    & All scenarios: worker-3 EWMA $\approx$190--200\,ms vs.\ $\approx$49\,ms. \\
R10 & The system \textbf{should} support deliberate worker crash for failure testing.
    & Should & POST \texttt{/api/v1/crash} returns; container stops
    & S4: worker-2 crashed at $t=20$\,s; marked unhealthy within 1\,s. \\
R11 & The load balancer \textbf{should} track per-worker EWMA latency for routing decisions.
    & Should & Inspect \texttt{ewma\_latency\_ms} field in \texttt{/lb/workers}
    & Tracked and verified across all scenarios; drives LA routing. \\
R12 & The system \textbf{may} expose a browser-accessible dashboard for live monitoring and test control.
    & Optional & Manual browser test at \texttt{/dashboard}
    & Dashboard deployed; accessible at \texttt{pcd.razc.ro}. \\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Conformance Checks}

Table~\ref{tab:conformance} lists the five conformance checks that
directly link requirements to experimental evidence. Each check
specifies a pass criterion derived from the measured results.

\begin{table}[h]
\caption{Conformance Checks}
\label{tab:conformance}
\centering
\small
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{@{}p{0.5cm}p{3.0cm}p{1.2cm}p{2.3cm}@{}}
\toprule
\textbf{ID} & \textbf{Check} & \textbf{Req.} & \textbf{Pass Criterion} \\
\midrule
C1 & Algorithm differentiation under constant load
   & R2, R3, R4
   & Worker-3 traffic: RR $\approx$33\%, LC $<$20\%, LA $<$10\% (S1 confirmed: 33.3\%, 13.4\%, 4.9\%). \\
C2 & Failure detection within one interval
   & R5, R6
   & Fewer than 2 failed requests after a worker crash at $t=20$\,s (S4: 1 failure). \\
C3 & Adaptive algorithms outperform RR under stress
   & R3, R4
   & Error rate $<$2\% and throughput $>$200\,req/s for LC and LA at 300\,VUs (S3: LC 1.63\%/231\,rps, LA 0.80\%/229\,rps; RR 7.59\%/173\,rps). \\
C4 & Prometheus metrics endpoint conforms to text format
   & R7
   & \texttt{curl /lb/metrics} produces valid Prometheus exposition with per-worker labels (verified; Grafana ingests successfully). \\
C5 & Experiment reproducibility
   & R8, R9
   & Running \texttt{bash scripts/run-experiments.sh} produces 37 result files covering all algorithm--scenario combinations without manual intervention. \\
\bottomrule
\end{tabular}
\end{table}

% ---------------------------------------------------------------
\section{Discussion}
% ---------------------------------------------------------------

\subsection{Key Findings}

\textbf{Adaptive algorithms consistently outperform round-robin.}
Across all four scenarios, both LC and LA reduce average latency,
increase throughput, and --- under stress --- cut failure rates by
5--10$\times$. The benefit is present even at moderate load (30\,VUs),
not only at saturation. This is because any traffic sent to the slow
worker imposes a latency penalty on the client that issued the request;
minimising that share directly reduces average latency.

\textbf{Latency-aware routing most aggressively avoids slow workers.}
By combining EWMA latency with in-flight count in a single score, LA
sends only 5--13\% of traffic to worker-3 across all scenarios, compared
to 13--17\% for LC and 33\% for RR. The EWMA score encodes both current
load (via in-flight) and historical speed (via EWMA), making it more
selective than in-flight count alone.

\textbf{Stress testing exposes correctness requirements.} At 300\,VUs,
round-robin's equal distribution overloads worker-3, whose slow
responses cause request timeouts, which cause health-check failures,
which cascade into all three workers being marked unhealthy. Adaptive
algorithms, by directing less traffic to worker-3, avoid this cascade.

\textbf{Failure recovery is near-instantaneous.} With a 1\,s health
check interval, the mean time to isolate a crashed worker is at most
1\,s. In our experiment, only one request was lost. After isolation,
the latency-aware algorithm correctly prefers the faster surviving
worker (worker-1 at $\sim$67\,ms over worker-3 at $\sim$215\,ms).

\textbf{Fairness and performance trade off against each other.}
Jain's fairness index $J = (\sum x_i)^2 / (n \sum x_i^2)$ measures
how evenly traffic is distributed across workers, with $J=1$ being
perfectly equal. Round-robin achieves $J=1.000$ by design, but this
fairness comes at the cost of routing to the slow worker at its full
equal share. Least-connections reduces fairness to $J=0.847$--$0.891$
across scenarios; latency-aware to $J=0.733$--$0.839$. The lower
fairness values directly correspond to better performance: fewer
requests reach the slow worker, reducing average latency and
increasing throughput. In heterogeneous deployments, equal traffic
distribution is not a goal --- proportional utilisation of actual
capacity is.

\textbf{Implementation correctness is as important as algorithm choice.}
The two bugs described in Section~V each individually caused all
algorithms to produce identical 33\% distributions. This result is
counter-intuitive: both are concurrency correctness issues that are
invisible in low-load unit tests but dominant at moderate concurrency.
They serve as a cautionary example that correctness and performance are
not separable concerns in async systems.

\subsection{Algorithm Trade-offs}

\textbf{Throughput vs.\ tail latency.} All algorithms exhibit high
p95 values at 300\,VUs ($>$2\,s) because the system is overloaded.
Adaptive algorithms improve average latency but cannot eliminate tail
latency once fast workers are also saturated.

\textbf{Computational overhead.} Round-robin is O(1) per request;
LC and LA are O($n$) over healthy workers. With $n=3$ the overhead
is negligible. At hundreds of backends, a min-heap would be required.

\textbf{EWMA convergence warmup.} With $\alpha=0.25$, roughly 10--15
requests are needed to converge from the 100\,ms initial value.
During this warmup period LA behaves closer to round-robin. Frequent
restarts or very short-lived services amplify this effect.

% ---------------------------------------------------------------
\section{Security, Privacy, Safety, and Ethical Considerations}
% ---------------------------------------------------------------

\subsection{Security}

The load balancer exposes several sensitive endpoints that must be
access-controlled in any deployment beyond a private network:
\begin{itemize}
    \item \texttt{POST /api/v1/crash} terminates a worker process. In
    production this endpoint must be removed or restricted to
    authenticated operators only, as unauthenticated access constitutes
    a trivial denial-of-service vector.
    \item \texttt{GET /lb/stats} and \texttt{GET /lb/workers} reveal
    internal topology (worker base URLs, error rates, EWMA latency
    values). This information aids an attacker in identifying degraded
    nodes or predicting routing behaviour for targeted traffic shaping.
    \item The test-runner service executes k6 as a subprocess. The
    scenario parameters (\texttt{vus}, \texttt{duration}) are passed
    to k6 via environment variables; input validation must ensure these
    cannot be used for command injection or resource exhaustion.
\end{itemize}

The current deployment at \texttt{pcd.razc.ro} is accessible over
HTTPS (via a wildcard Let's Encrypt certificate) but has no
authentication layer. This is acceptable for a course prototype but
would require at minimum HTTP Basic Auth or OAuth before any production use.

\subsection{Privacy}

The system processes synthetic analytics events with no real user
data. Prometheus stores metric time-series (request counts, latencies,
worker health) which are aggregated and contain no personally
identifiable information. Log files record HTTP request paths and
source IP addresses; in a production system these would require data
retention policies in compliance with applicable regulations (e.g., GDPR).

\subsection{Safety}

The \texttt{/api/v1/crash} endpoint is intentionally destructive.
In a real infrastructure deployment, calling this endpoint on a
production worker would cause service disruption. The endpoint is
useful for controlled chaos engineering but must be gated behind
a change-management process (e.g., maintenance window, approval
workflow) in production environments.

\subsection{Ethical Considerations}

All load tests were conducted on infrastructure owned and operated
by the authors. No public services or third-party systems were targeted.
The k6 load generator and all dependent services are open-source.
The experimental results and implementation are made publicly available
to support reproducibility and future research.

% ---------------------------------------------------------------
\section{Limitations and Threats to Validity}
% ---------------------------------------------------------------

\subsection{Internal Validity}

\textbf{Single-host confounding.} All containers share the same 4-core
CPU and memory. At high VU counts, CPU scheduling contention between
the k6 process, the load balancer, and three workers may alter results
in ways that would not occur on a multi-host deployment. The relative
ordering of algorithms is unlikely to change, but absolute latency and
throughput values are hardware-specific.

\textbf{Short experiment duration.} Each run lasts 60--70\,s. EWMA
convergence, long-term steady-state distributions, and recovery from
repeated transient failures cannot be studied at this timescale.

\textbf{Single repetition.} Each scenario was run once per algorithm.
Without repeated runs we cannot report confidence intervals or test
for statistical significance. Performance variability between runs on
shared hardware may be non-negligible.

\textbf{Synthetic workload.} The \texttt{asyncio.sleep} + busy-loop
model produces deterministic service times without the queue-length
feedback or memory effects present in real analytics backends
(e.g., database query plans, JVM JIT warmup, OS page cache effects).

\subsection{External Validity}

\textbf{Generalisation to larger worker pools.} With $n=3$ workers,
the routing decision space is small. With $n=100$, the relative
advantage of EWMA scoring over in-flight counting may change because
there are more equally-fast alternatives to route to.

\textbf{Generalisation to heterogeneous workloads.} All requests carry
the same \texttt{delay\_ms} and \texttt{cpu\_ms} parameters. In
production analytics systems, request cost is often highly skewed
(e.g., cold-start vs.\ cached queries), which would favour adaptive
algorithms even more strongly.

\textbf{Fixed EWMA $\alpha$.} The decay factor $\alpha=0.25$ was
chosen empirically. A systematic sweep across $\alpha$ values would
be needed to characterise the convergence/stability trade-off and
determine whether a different value would change the ranking of algorithms.

\subsection{Construct Validity}

\textbf{EWMA at end of test as proxy for routing behaviour.}
The worker-level EWMA values reported in the results tables are
captured at the end of each experiment and reflect the steady-state
estimate, not the transient during the first 10--15 requests. During
warmup, LA behaves closer to RR.

\textbf{Rolling latency window.} The p95/p99 values in
\texttt{/lb/stats} cover only the last 1000 requests. For the stress
test, where conditions change rapidly, this window reflects the final
load level rather than the worst-seen during the ramp.

% ---------------------------------------------------------------
\section{Conclusion and Future Work}
% ---------------------------------------------------------------

We designed, implemented, and evaluated a containerised load-balancing
testbed comparing round-robin, least-connections, and latency-aware
EWMA-based routing across four controlled workload scenarios. The
principal findings are:

\begin{enumerate}
    \item \textbf{Adaptive algorithms outperform round-robin across all
    conditions.} Latency-aware routing reduces average latency by 30\%
    and increases throughput by 19\% at constant moderate load. Under
    extreme stress (300\,VUs), both adaptive algorithms sustain 32\%
    higher throughput with 5--10$\times$ lower error rates.
    \item \textbf{Failure recovery is near-instantaneous} with a 1\,s
    health-check interval: the system loses only one request out of 9252
    after a complete worker crash.
    \item \textbf{Implementation correctness is critical.} Two concurrency
    bugs --- EWMA contamination by health-check latency and in-flight
    counter increment after an asyncio yield point --- caused all
    algorithms to produce identical routing distributions until fixed.
\end{enumerate}

\textbf{Future work} directions include:
\begin{itemize}
    \item \textbf{Shorter health-check intervals and circuit breakers.}
    A 200--500\,ms health-check window combined with a circuit-breaker
    pattern (trip after $k$ consecutive failures) would reduce the
    failure-exposure window further.
    \item \textbf{Thread-pool isolation for CPU-bound work.} Moving
    the synchronous busy-loop to \texttt{asyncio.run\_in\_executor}
    would decouple request processing from health-check responsiveness,
    making worker-3 a more realistic model of a slow but healthy backend.
    \item \textbf{Reinforcement-learning routing.} An RL agent that
    treats each request as an action could learn non-stationary
    workload patterns without explicit EWMA parameter tuning.
    \item \textbf{Systematic $\alpha$ sweep.} Evaluating EWMA decay
    factors across $\alpha \in \{0.1, 0.25, 0.5, 0.75\}$ would
    quantify the convergence--stability trade-off.
    \item \textbf{Statistical replication.} Running each scenario five
    or more times and reporting means with 95\% confidence intervals
    would strengthen the validity of the conclusions.
\end{itemize}

\begin{thebibliography}{10}

\bibitem{nginx}
NGINX Inc., ``NGINX Load Balancing,''
\url{https://docs.nginx.com/nginx/admin-guide/load-balancer/http-load-balancer/},
accessed 2024.

\bibitem{haproxy}
HAProxy Technologies, ``HAProxy Configuration Manual,''
\url{https://docs.haproxy.org/}, accessed 2024.

\bibitem{kubernetes}
The Kubernetes Authors, ``Service -- Kubernetes Documentation,''
\url{https://kubernetes.io/docs/concepts/services-networking/service/},
accessed 2024.

\bibitem{tanenbaum}
A.~S. Tanenbaum and M.~Van Steen,
\textit{Distributed Systems: Principles and Paradigms}, 3rd ed.
Pearson, 2017.

\bibitem{cardellini}
V.~Cardellini, M.~Colajanni, and P.~S. Yu,
``Dynamic load balancing on Web-server systems,''
\textit{IEEE Internet Computing}, vol.~3, no.~3, pp.~28--39, 1999.

\bibitem{pai1998}
V.~S. Pai, M.~Aron, G.~Banga, M.~Svendsen, P.~Druschel,
W.~Zwaenepoel, and E.~Nahum,
``Locality-aware request distribution in cluster-based network servers,''
in \textit{Proc. ASPLOS}, 1998, pp.~205--216.

\bibitem{envoy}
Envoy Proxy, ``Load Balancing Policies,''
\url{https://www.envoyproxy.io/docs/envoy/latest/intro/arch_overview/upstream/load_balancing/load_balancers},
accessed 2024.

\bibitem{mitzenmacher2001}
M.~Mitzenmacher,
``The power of two choices in randomized load balancing,''
\textit{IEEE Trans. Parallel Distrib. Syst.}, vol.~12, no.~10,
pp.~1094--1104, 2001.

\bibitem{jacobson1988}
V.~Jacobson,
``Congestion avoidance and control,''
\textit{ACM SIGCOMM Comput. Commun. Rev.}, vol.~18, no.~4,
pp.~314--329, 1988.

\bibitem{docker}
Docker Inc., ``Docker Compose Overview,''
\url{https://docs.docker.com/compose/}, accessed 2024.

\end{thebibliography}

\end{document}
